### 2023-01-22 doc/src/MSLevel.tex
Allow mvien + mvip virtual interrupts SEI, SSI
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

When supervisor mode is implemented, the Advanced Interrupt
Architecture adds a facility for software filtering
of interrupts and for virtual interrupts, making use of new
CSRs \z{mvien}/\z{mvienh} (Machine Virtual Interrupt Enables) and
\z{mvip}/\z{mviph} (Machine Virtual Interrupt-Pending bits).
\emph{Interrupt filtering} permits a local or custom interrupt
to trap to \mbox{M-mode} and then be selectively delegated by
software to supervisor level, even while the corresponding bit in
\z{mideleg}/\z{midelegh} remains zero.
The same hardware may also, under the right circumstances, allow
machine level to assert \emph{virtual interrupts} to supervisor level
that have no connection to any real interrupt events.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c||c|c|}
\hline
\z{mideleg[}$n$\z{]} & \z{mvien[}$n$\z{]} &
    \z{sip[}$n$\z{]} & \z{sie[}$n$\z{]} \\
\hline
\hline
0 & 0  & Read-only 0                & Read-only 0 \\
0 & 1  & Alias of \z{mvip[}$n$\z{]} & Writable \\
1 & -- & Alias of \z{mip[}$n$\z{]}  & Alias of \z{mie[}$n$\z{]} \\
\hline
\end{tabular}
\end{center}
\caption{%
The effects of \z{mideleg} and \z{mvien} on \z{sip} and \z{sie} (except
for the hypervisor extension's VS-level interrupts, which appear in
\z{hip} and \z{hie} instead of \z{sip} and \z{sie}).
A bit in \z{mvien} can be set to~1 only for major interrupts 13--63.
When interrupts 0--12 are delegated by \z{mideleg}, the
aliases in \z{sip} may be read-only, as specified by the
{\RISCV} Privileged Architecture.%
}
\label{tab:intrFilteringForS}
\end{table*}

@@

\begin{commentary}
Platform standards may require that interrupt filtering be supported
for certain interrupt causes.
\end{commentary}

Bits 12:0 in \z{mvien} are reserved and must be read-only zeros.
Bits 1 and~5 of \z{mvip} are aliases of the same bits in \z{mip}
(SSIP and STIP) when those bits are writable in \z{mip}.
When STIP is not writable in \z{mip} (such as when optional
CSR \z{stimecmp} is implemented), the
same bit in \z{mvip} is read-only zero.
Bit~9 of \z{mvip} is an alias of the software-writable bit~9 of \z{mip}
(SEIP).
The other bits of \z{mvip} in the range 12:0 are reserved and must be
read-only zeros.

\begin{commentary}
Bit~9 of\/ \z{mvip} makes the software-writable SEIP bit of\/ \z{mip}
directly accessible by itself.
\end{commentary}

@@

When supervisor mode is implemented, the minimal required
implementation of \z{mvien}/\z{mvienh} and \z{mvip}/\z{mviph} has all
bits being read-only zeros except for \z{mvip} bits 1, 5, and~9, which
are aliases of existing writable bits in \z{mip}.
When supervisor mode is not implemented, registers \z{mvien}/\z{mvienh}
and \z{mvip}/\z{mviph} do not exist.

>>>>

When supervisor mode is implemented, the Advanced Interrupt
Architecture adds a facility for software filtering
of interrupts and for virtual interrupts, making use of new
CSRs \z{mvien}/\z{mvienh} (Machine Virtual Interrupt Enables) and
\z{mvip}/\z{mviph} (Machine Virtual Interrupt-Pending bits).
\emph{Interrupt filtering} permits a supervisor-level
interrupt (SEI or SSI) or local or custom interrupt
to trap to \mbox{M-mode} and then be selectively delegated by
software to supervisor level, even while the corresponding bit in
\z{mideleg}/\z{midelegh} remains zero.
The same hardware may also, under the right circumstances, allow
machine level to assert \emph{virtual interrupts} to supervisor level
that have no connection to any real interrupt events.

@@

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|c||c|c|}
\hline
\z{mideleg[}$n$\z{]} & \z{mvien[}$n$\z{]} &
    \z{sip[}$n$\z{]} & \z{sie[}$n$\z{]} \\
\hline
\hline
0 & 0  & Read-only 0                & Read-only 0 \\
0 & 1  & Alias of \z{mvip[}$n$\z{]} & Writable \\
1 & -- & Alias of \z{mip[}$n$\z{]}  & Alias of \z{mie[}$n$\z{]} \\
\hline
\end{tabular}
\end{center}
\caption{%
The effects of \z{mideleg} and \z{mvien} on \z{sip} and \z{sie} (except
for the hypervisor extension's VS-level interrupts, which appear in
\z{hip} and \z{hie} instead of \z{sip} and \z{sie}).
A bit in \z{mvien} can be set to~1 only
for major interrupts 1, 9, and 13--63.
For interrupts 0--12,
aliases in \z{sip} may be read-only, as specified by the
{\RISCV} Privileged Architecture.%
}
\label{tab:intrFilteringForS}
\end{table*}

@@

\begin{commentary}
Platform standards or other extensions may require that
bits of\/ \z{mvien} for certain interrupt causes
be writable, or be read-only zero or one.
\end{commentary}

The bits of \z{mvien} for supervisor software interrupts (code~1)
and supervisor external interrupts (code~9)
are each either writable or read-only zero;
they cannot be read-only ones.
The other bits of \z{mvien} for interrupts 0--12
are reserved and must be read-only zeros.

It is strongly recommended that bit~9 of \z{mvien} be writable.
Furthermore, if bit~1 (SSIP) of \z{mip} can be set
automatically by an interrupt controller and
not just by explicit writes to \z{mip} or \z{sip},
it is strongly recommended that bit~1 of \z{mvien} also be writable.

When bit~1 of \z{mvien} is zero, bit~1 of \z{mvip}
is an alias of the same bit (SSIP) of \z{mip}.
But when bit~1 of \z{mvien} is one, bit~1 of \z{mvip}
is a separate writable bit independent of \z{mip}.SSIP.
When the value of bit~1 of \z{mvien} is changed from zero to one,
the value of bit~1 of \z{mvip} becomes {\unspecified}.

Bit~5 of \z{mvip} is an alias of the same bit (STIP)
in \z{mip} when that bit is writable in \z{mip}.
When STIP is not writable in \z{mip} (such as when
\z{menvcfg}.STCE =~1), bit~5 of \z{mvip} is read-only zero.

When bit~9 of \z{mvien} is zero, bit~9 of \z{mvip} is
an alias of the software-writable bit~9 of \z{mip} (SEIP).
But when bit~9 of \z{mvien} is one, bit~9 of \z{mvip}
is a writable bit independent of \z{mip}.SEIP\@.
Unlike for bit~1, changing the value of bit~9 of \z{mvien}
does not affect the value of bit~9 of \z{mvip}.

\begin{commentary}
When bit~9 of\/ \z{mvien} is zero, bit~9 of\/ \z{mvip}
makes the software-writable SEIP bit of\/ \z{mip}
directly accessible by itself.
\end{commentary}

Except for bits 1, 5, and~9 as specified above, the bits of \z{mvip}
in the range 12:0 are reserved and must be read-only zeros.

The value of bit~9 of \z{mvien} has some additional consequences
for supervisor external interrupts:
\begin{itemize}

\item
When bit~9 of \z{mvien} is zero, the software-writable SEIP bit
(bit~9 of \z{mvip}) interacts with reads and writes of \z{mip}
in the way specified by the {\RISCV} Privileged Architecture.
In particular, for most purposes, the value of bit~9 of \z{mvip}
is logically ORed into the readable value of \z{mip}.SEIP\@.
But when bit~9 of \z{mvien} is one, bit SEIP in \z{mip} is
read-only and does not include the value of bit~9 of \z{mvip}.
Rather, the value of \z{mip}.SEIP is simply
the supervisor external interrupt signal from
the hart's external interrupt controller (APLIC or IMSIC).

\item
If the hart has an IMSIC, then when bit~9 of \z{mvien} is one,
attempts from S-mode to explicitly access the supervisor-level
interrupt file raise an illegal instruction exception.
The exception is raised for attempts to access CSR \z{stopei},
or to access \z{sireg} when \z{siselect} has
a value in the range \z{0x70}--\z{0xFF}.
Accesses to guest interrupt files
(through \z{vstopei} or \z{vsiselect} + \z{vsireg}) are not affected.

@@

When supervisor mode is implemented, the minimal required
implementation of \z{mvien}/\z{mvienh} and \z{mvip}/\z{mviph} has all
bits being read-only zeros except for \z{mvip}
bits 1 and~9, and sometimes bit~5, each of which is
an alias of an existing writable bit in \z{mip}.
(Although, as noted, it is strongly recommended
that bit 9 of \z{mvien} also be writable.)
When supervisor mode is not implemented, registers \z{mvien}/\z{mvienh}
and \z{mvip}/\z{mviph} do not exist.
